---
title: "Practical Machine Learning"
author: "Ignacio Rodriguez Solis"
date: "5/11/2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background

Using devices such as Jawbone Up, Nike FuelBand and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, the goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways: A: exactly according to the specification; B: throwing the elbows to the front; C: lifting the dumbbell only halfway; D: lowering the dumbbell only halfway; E: throwing the hips to the front.

Loading the R packages needed for analysis:

```{r message=FALSE, warning=FALSE}
library(caret)
library(corrplot)
library(Rtsne)
library(xgboost)
library(stats)
library(knitr)
library(ggplot2)
library(rattle)
library(rpart)
library(randomForest)
library(repmis)
```


## Data

### Training Data

The training data for this project are available here:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

```{r cache=TRUE}
TrainingURL <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
TrainingFile = "./data/pml-training.csv"
download.file(TrainingURL, destfile = TrainingFile, method="curl")
TrainingData <- read.csv(file = "./data/pml-training.csv", header = TRUE, as.is = TRUE, stringsAsFactors = FALSE, sep = ',', na.strings = c('NA','','#DIV/0!'))
# summary(TrainingData)

```

The training data has 19622 rows of observations and 160 features (predictors). The testing data has 20 rows and the 160 features as well. There is one column of target outcome which is the classe variable in the training set.

### Testing Data

The testing data are available here:
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

```{r cache=TRUE}
TestingURL <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
TestingFile = "./data/pml-testing.csv"
download.file(TestingURL, destfile = TestingFile, method="curl")
TestingData <- read.csv(file = "./data/pml-testing.csv", header = TRUE, as.is = TRUE, stringsAsFactors = FALSE, sep = ',', na.strings = c('NA','','#DIV/0!'))
# summary(TestingData)
```

### Processing Training Data

There are a lot of NA values, useless or empty variables for the prediction. Besides, the prediction will be based on data from accelerometers on the belt, forearm, arm, and dumbell of the participants. Therefore, non-accelerometer measures will be disregarded as well:

```{r cache=TRUE}
TrainingData$classe <- as.factor(TrainingData$classe)

# Remove non-variance variables
ZeroVariance <- nearZeroVar(TrainingData)
TrainingData <- TrainingData[, -ZeroVariance]

# Remove NA-value variables
NAValues <- sapply(TrainingData, function(x) mean(is.na(x))) > 0.97
TrainingData <- TrainingData[, NAValues == FALSE]

# Remove variables non essential for the analysis and predictions, (columns one to six)
TrainingData <- TrainingData[, -(1:6)]

dim(TrainingData)
names(TrainingData)

# Cross Validation
set.seed(77)
SetTrainingData <- createDataPartition(y = TrainingData$classe, p = 0.7, list = FALSE)
TrainingData1 <- TrainingData[SetTrainingData, ]
ValidationData <- TrainingData[-SetTrainingData, ]
```

### Training Data Modelling

The model will be built based on random forest approach, due to its highly accuracy rate. Cross validation will be used as train control method:

```{r cache=TRUE}
# Random Forest
set.seed(77)
ModelControl1 <- trainControl(method = "cv", number = 3, verboseIter = FALSE)
TrainingModelFit1 <- train(classe ~ ., data = TrainingData1, method = "rf", trControl = ModelControl1)

TrainingModelFit1$finalModel
```

### Model Evaluation

I use the previous model on the training data, to predict the classe in ValidationData. A confusion matrix will be shown as well, to compare predicted versus actual data:

```{r cache=TRUE}
# Model to predict classe in ValidationData
PredictionModel1 <- predict(TrainingModelFit1, newdata = ValidationData)

# Confusion matrix to get estimated out-of-sample error
confusionMatrix(ValidationData$classe, PredictionModel1)
```

### Processing Testing Data

Every processing made to the Training Data is going to be repeated for the Testing Data as well. Besides, the model fit will be set now for the entire training dataset, and it will be used afterwards to predict the outcomes on the testing dataset:

```{r cache=TRUE}
# Remove non-variance variables
ZeroVariance <- nearZeroVar(TestingData)
TestingData <- TestingData[, -ZeroVariance]

# Remove NA-value variables
NAValues <- sapply(TestingData, function(x) mean(is.na(x))) > 0.97
TestingData <- TestingData[, NAValues == FALSE]

# Remove variables non essential for the analysis and predictions, (columns one to six)
TestingData <- TestingData[, -(1:6)]

dim(TestingData)
names(TestingData)

# Fit model again using full training dataset (TrainingData)
ModelControl <- trainControl(method = "cv", number = 3, verboseIter = FALSE)
TrainingModelFit <- train(classe ~ ., data = TrainingData, method = "rf", trControl = ModelControl)

```

### Predicting Testing Dataset

The model previously fitted for the training dataset will be now used to make "classe" predictions on the testing dataset:

```{r cache=TRUE}
# Model to predict classe in TestingData
TestingPredictionModel <- predict(TrainingModelFit, newdata = TestingData)

# Convert predictions in TestingData to characters
TestingPredictionModel <- as.character(TestingPredictionModel)

TestingPredictionModel
```
